Tools for identifying unexpectedly low microfossil count sums

### 2020-08-19 Rmd/count_check_MS.Rmd
improved abstract - methods

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported count sums, are identified.
journal: "Journal of Quaternary Science"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

@@

If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts.
Unfortunately, count data might be falsified so that it appears to meet the reported count sum.
A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum.
This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than reported.

The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e. few assemblages should have all counts divisible by an integer larger than one, especially if species richness is high.

This paper aims to test the methods presented above and present some cases with unexpected results.
Some complications and caveats are discussed.

@@

For count data, the prevalence of counts without singletons and the 
greatest common divisor of each count were calculated and related to the count sum.
For percent data, the count sum was estimated using the minimum percent method and the direct search methods described above.

All analyses were done in R v. `r getRversion()` [@R] and used the packages 
extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
numbers v. `r packageVersion("numbers")` [@numbers], 
countSum v. `r packageVersion("countSum")` [@countSum], 
neotoma v. `r packageVersion("neotoma")` [@neotoma], 
drake v. `r packageVersion("drake")` [@drake], 
and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
journal: "Journal of Quaternary Science"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

@@

Given that count sums of percent data have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
This paper aim to develop and test simple methods that can flag if the count data might have been misreported, or if the count sum of percent data is perhaps lower than reported.

The key insight that allows inference about community or assemblage count data is that they represent a species abundance distribution.
The ubiquitous shape of species abundance distributions when plotted as as the number of species against abundance is a hyperbolic or hollow curve with many rare and few abundant species [@McGill_2007].
This relationship is central to ecological theory. 
Communities or assemblages data that differ markedly this expected relationship

For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].

In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e. few assemblages should have all counts divisible by an integer larger than one, especially if species richness is high.

Some complications and caveats are discussed.

<!-- If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts. -->
<!-- Unfortunately, count data might be falsified so that it appears to meet the reported count sum. -->
<!-- A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum. -->

@@

For count data, the prevalence of counts without singletons and the 
greatest common divisor of each count were calculated and related to the count sum.

Generalised additive model proportion of explained deviance
Proportion of taxa divisible by _k_ where $k > 1$

All analyses were done in R v. `r getRversion()` [@R] and used the packages 
extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
numbers v. `r packageVersion("numbers")` [@numbers], 
countSum v. `r packageVersion("countSum")` [@countSum], 
neotoma v. `r packageVersion("neotoma")` [@neotoma], 
and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Packages drake v. `r packageVersion("drake")` [@drake], 
bookdown `r packageVersion("bookdown")` [@bookdown] and
renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
